---
title: "s3-4 Evaluation of ERCC control"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
     fig_width: 5
     fig_height: 5
---


# Overview

ERCC spike-in RNA was added to polysome fractionated samples.
Library normalization factors will be calculated based on the reads counts of ERCC RNAs.

```{r load_libraries, message = FALSE, warning = FALSE}

## Bioconductor
library("DESeq2")

## Specify the number of CPUs to be used
processors <- 8

temp <- sapply(list.files("../functions", full.names = TRUE), source)

```


```{r define_directory}

sample.file <- file.path("../../data/sample_data/processed_sample_file.csv")

## Input annotation
annot.dir <- file.path("../../annotation/")

annot.ps.dir <- file.path(annot.dir, "hg38_annotation/processed_data/")
annot.R.file <- list.files(
    annot.ps.dir,
    pattern = glob2rx("*primary_transcript_annotation*.rdata"),
    full.names = TRUE
)
load(annot.R.file)

## Input files
results.dir <- file.path("../../results")
s3.alignment.stats.dir <- file.path(results.dir, "s3-alignment-statistics")
s3.4.poly.size.factor.dir <- file.path(s3.alignment.stats.dir, "polysome_size_factor")

create.dirs(c(s3.4.poly.size.factor.dir))

```


```{r read_start_count_data}

sample.dt <- fread(sample.file)
sample.names <- sample.dt[, sample_name]

ercc.count.dt <- file.path(
    s3.alignment.stats.dir,
    "ERCC_count_per_sample.csv"
) %>% fread

```



# Calculate sum of reads from ERCC RNAs


```{r ercc_count_and_each_fraction}

## Sample with ERCC data are selected here
sl.sample.names <- sample.dt[
    ERCC == "Yes",
    sample_name
]

m.ercc.count.dt <- melt(
    ercc.count.dt,
    id.vars = "ERCC_id",
    variable.name = "sample_name",
    value.name = "count"
)

m.ercc.count.dt[
  , seqgroup := ifelse(grepl("^ERCC", ERCC_id), "ERCCs", ERCC_id)
]

ercc.sum.count.dt <- m.ercc.count.dt[
  , .(sum_count = sum(count)), by = list(sample_name, seqgroup)
]

d.ercc.sum.count.dt <- dcast(
    ercc.sum.count.dt,
    sample_name ~ seqgroup,
    value.var = "sum_count"
)

```


# Excluding ribo 0A and 0B fraction

This is a default procedure.

## Calculate library size factor based on read counts of ERCC RNAs


```{r calculate_normalization_factor_based_on_ERCC}

non0.sl.sample.names <- sl.sample.names[!grepl("(ribo0A|ribo0B)", sl.sample.names)]

dds <- DESeqDataSetFromMatrix(
    countData = ercc.count.dt[
        grepl("^ERCC-", ERCC_id),
        non0.sl.sample.names,
        with = FALSE
    ],
    colData = sample.dt[
        sample_name %in% non0.sl.sample.names
    ],
    design = ~ 1
) %>%
    estimateSizeFactors

sizefactor.dt <- sizeFactors(dds) %>%
    data.frame %>%
    data.table(keep.rownames = TRUE)

setnames(
    sizefactor.dt,
    old = c("rn", "."),
    new = c("sample_name", "size_factor")
)

sample.count.with.sizefactor.dt <- merge(
    sample.dt,
    sizefactor.dt,
    by = "sample_name",
    all = TRUE
) %>%
    merge(
        y = d.ercc.sum.count.dt,
        by = "sample_name",
        all = TRUE
    )

```


## Comparison of DESeq2 calculated size factors and total read counts from ERCC RNAs


The analysis below indicates that the choice between using DESeq2 calculated size factors and total read counts of ERCC spike-in RNAs does not have a significant impact on the downstream analyses.
Thus, I decided to use the DESeq2 calculated size factors.


```{r ERCC_total_vs_DESeq2_size_factors}

ercc.total.size.factor.lm <- lm(
    ERCCs ~ size_factor,
    data = sample.count.with.sizefactor.dt[sample_name %in% non0.sl.sample.names]
)

sample.count.with.sizefactor.dt[
    sample_name %in% non0.sl.sample.names,
    residual2lm_ERCC_sizefactor := resid(ercc.total.size.factor.lm)
]

poly.sample.count.with.sizefactor.dt <- sample.count.with.sizefactor.dt[
    grepl("^ribo", fraction)
]

poly.sample.count.with.sizefactor.dt[, `:=`(
    size_factor_by_count_sum = ERCCs / mean(ERCCs)
)]

ggplot(
    data = poly.sample.count.with.sizefactor.dt,
    aes(
        x = ERCCs,
        y = size_factor,
        color = TSO
    )
) +
    geom_point() +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    )

ggplot(
    data = poly.sample.count.with.sizefactor.dt,
    aes(
        x = size_factor_by_count_sum,
        y = size_factor,
        color = TSO
    )
) +
    geom_point() +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    )


```



## Export size factors


```{r export_predicted_size_factors}

poly.sample.count.with.sizefactor.dt[
   , .(sample_name, size_factor, size_factor_by_count_sum)
] %>%
    fwrite(
        file = file.path(
            s3.4.poly.size.factor.dir,
            "library_size_factor_by_ERCC.csv"
            )
    )


```


# Including ribo 0A and 0B fraction


## Calculate library size factor based on read counts of ERCC RNAs


```{r calculate_normalization_factor_based_on_ERCC_with_ribo_0}

dds <- DESeqDataSetFromMatrix(
    countData = ercc.count.dt[
        grepl("^ERCC-", ERCC_id),
        sl.sample.names,
        with = FALSE
    ],
    colData = sample.dt[
        sample_name %in% sl.sample.names
    ],
    design = ~ 1
) %>%
    estimateSizeFactors

sizefactor.dt <- sizeFactors(dds) %>%
    data.frame %>%
    data.table(keep.rownames = TRUE)

setnames(
    sizefactor.dt,
    old = c("rn", "."),
    new = c("sample_name", "size_factor")
)

sample.count.with.sizefactor.dt <- merge(
    sample.dt,
    sizefactor.dt,
    by = "sample_name",
    all = TRUE
) %>%
    merge(
        y = d.ercc.sum.count.dt,
        by = "sample_name",
        all = TRUE
    )

```


## Comparison of DESeq2 calculated size factors and total read counts from ERCC RNAs


The analysis below indicates that the choice between using DESeq2 calculated size factors and total read counts of ERCC spike-in RNAs does not have a significant impact on the downstream analyses.
Thus, I decided to use the DESeq2 calculated size factors.


```{r ERCC_total_vs_DESeq2_size_factors_with_ribo_0}

ercc.total.size.factor.lm <- lm(
    ERCCs ~ size_factor,
    data = sample.count.with.sizefactor.dt[sample_name %in% sl.sample.names]
)

sample.count.with.sizefactor.dt[
    sample_name %in% sl.sample.names,
    residual2lm_ERCC_sizefactor := resid(ercc.total.size.factor.lm)
]

poly.sample.count.with.sizefactor.dt <- sample.count.with.sizefactor.dt[
    grepl("^ribo", fraction)
]

poly.sample.count.with.sizefactor.dt[, `:=`(
    size_factor_by_count_sum = ERCCs / mean(ERCCs)
)]

ggplot(
    data = poly.sample.count.with.sizefactor.dt,
    aes(
        x = ERCCs,
        y = size_factor,
        color = TSO
    )
) +
    geom_point() +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    )

ggplot(
    data = poly.sample.count.with.sizefactor.dt,
    aes(
        x = size_factor_by_count_sum,
        y = size_factor,
        color = TSO
    )
) +
    geom_point() +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    )


```



## Export size factors with ribo0


```{r export_predicted_size_factors_with_ribo_0}

poly.sample.count.with.sizefactor.dt[
   , .(sample_name, size_factor, size_factor_by_count_sum)
] %>%
    fwrite(
        file = file.path(
            s3.4.poly.size.factor.dir,
            "library_size_factor_with_ribo0_by_ERCC.csv"
            )
    )


```



# Session information


```{r sessionInfo}

sessionInfo()

```



