---
title: "s9-1-4 Interplay of the HIF and mTOR pathways"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   html_document:
     highlight: haddock
     toc: yes
     toc_depth: 2
     keep_md: yes
     fig_width: 5
     fig_height: 5
---


# Overview


```{r load libraries, message = FALSE, warning = FALSE}

sig.th <- 0.05

## Data visualization packages
library("ggrepel")
library("cowplot")
## Library("kableExtra")
## Parallelization oackages
## Specify the number of CPUs to be used
processors <- 8

temp <- sapply(list.files("../functions", full.names = TRUE), source)
temp <- sapply(list.files("../s8-analysis-of-translation/functions", full.names = TRUE), source, chdir = TRUE)
temp <- sapply(list.files("../s9-integrative-analysis/functions", full.names = TRUE), source, chdir = TRUE)
source("../s6-differential-expression-and-tss-usage/functions/load_total_analysis_results.R", chdir = TRUE)

set.seed(0)

```


```{r define directory}

results.dir <- file.path("../../results")
s7.dir <- file.path(results.dir, "s7-HIF-binding-site")
s7.2.hif1.2a.dir <- file.path(s7.dir, "s7-2-HIF1A-and-HIF2A-ratio")

s8.dir <- file.path(results.dir, "s8-analysis-of-translation")
s8.1.dir <- file.path(s8.dir, "s8-1-differentially-translated-mRNAs")
s8.1.1.dir <- file.path(s8.1.dir, "gene-level-dte")
s8.3.dir <- file.path(s8.dir, "s8-3-validation-of-method")

s9.dir <- file.path(results.dir, "s9-integrative-analysis")

sample.file <- file.path("../../data/sample_data/processed_sample_file.csv")
sample.dt <- fread(sample.file)
sample.names <- sample.dt[, sample_name]

```

# Pre-processed data import


```{r load preprocessed data}

all.de.dte.res.dt <- fread(
    file.path(
        s9.dir,
        "all-differential-expression-and-translation-data.csv"
    ),
    na.strings = ""
)

all.de.dte.res.dt[, `:=`(
    VHL_target_RCC4 = factor(
        VHL_target_RCC4,
        levels = c("VHL_loss_repressed", "non_VHL_target", "VHL_loss_induced")
    ),
    mRNA_trsl_intersection_by_functions =  factor(
        mRNA_trsl_intersection_by_functions,
        levels = c("Glycolysis", "Others", "Angiogenesis or vascular process")
    )
)]

all.filtered.gene.dt <- fread(
    file.path(
        s8.3.dir,
        "filtered_gene_for_polysome_analysis.csv"
    )
)

th.all.de.dte.res.dt <- all.de.dte.res.dt[
        gene_id %in% intersect(
                         all.filtered.gene.dt[RCC4_noVHL_NA == TRUE, gene_id],
                         all.filtered.gene.dt[RCC4_noVHL_Torin1== TRUE, gene_id]
                     )
]


```


# GO term enrichment analysis of HIF target genes by their changes in translation upon mTOR inhibition


```{r analyze GO term enrichment, fig.height = 5, fig.width = 6}

library("gprofiler2")
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15") # For reproducibility, use archive version

print("The number of genes for GO term analysis")
th.all.de.dte.res.dt[, table(RCC4_mRNA_mTOR_trsl_group)]
print(paste0("Total number of genes in the background: ", nrow(th.all.de.dte.res.dt)))

gostres.hif.up.mtor.preserved <- gost(
    query = th.all.de.dte.res.dt[
        RCC4_mRNA_mTOR_trsl_group == "mRNA_up_and_translation_up",
        gene_id
    ],
    custom_bg = th.all.de.dte.res.dt[, gene_id],
    organism = "hsapiens"
)

gostres.hif.up.mtor.preserved.res.dt <-
    gostres.hif.up.mtor.preserved$result %>% data.table


gostres.hif.up.mtor.down <- gost(
    query = th.all.de.dte.res.dt[
        RCC4_mRNA_mTOR_trsl_group == "mRNA_up_and_translation_down",
        gene_id
    ],
    custom_bg = th.all.de.dte.res.dt[, gene_id],
    organism = "hsapiens"
)

gostres.hif.up.mtor.down.res.dt <-
    gostres.hif.up.mtor.down$result %>% data.table


mrna.mtor.go.res.dt <- rbindlist(list(
    gostres.hif.up.mtor.preserved.res.dt[
      , regulation_type := "VHL_loss_induced_and_translation_preserved"],
    gostres.hif.up.mtor.down.res.dt[
      , regulation_type := "VHL_loss_induced_and_translation_down-regulated"]
))

sl.mrna.mtor.go.res.dt <- mrna.mtor.go.res.dt[order(p_value)] %>%
    {.[!(source %in% c("MIRNA", "WP", "HP"))]} %>%
    {.[, head(.SD, 5), by = regulation_type]}

sl.mrna.mtor.go.res.dt[order(regulation_type), .(
    regulation_type,
    term_name, source, p_value 
)] %>% print

file.path(
    s9.dir, "VHL-mTOR-GO-term-analysis-result.csv"
) %>% {fwrite(sl.mrna.mtor.go.res.dt, .)}

```


# HIF and mTOR dependent regulation of glycolytic genes


```{r hif and mTOR on glycolytic genes}

to.show.th.dt <- copy(th.all.de.dte.res.dt)

to.show.th.dt <- to.show.th.dt[
    (mRNA_trsl_intersection_by_functions == "Glycolysis") | gene_name %in% c("TYMP", "PGM2", "DERA")
][, .(gene_name,
      ##RCC4_mRNA_mTOR_trsl_group,
      translational_regulation_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
      VHL_target_RCC4,
      log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
      padj_RCC4_xx_HIF1B_N__noVHL_vs_VHL
      )
  ][order(gene_name)]

setnames(
    to.show.th.dt,
    old = c(
        "VHL_target_RCC4",
        "translational_regulation_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA",
        "log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL",
        "padj_RCC4_xx_HIF1B_N__noVHL_vs_VHL"),
    new = c("mRNA_by_VHL_loss", "mTOR_inh_noVHL", "mRNA_log2fc", "mRNA_padj")
)

to.show.th.dt[
    order(match(
        gene_name,
        c(
            "HK1", "GPI", "PFKL", "PFKM", "ALDOA", "ALDOC",
            "GAPDH", "TPI1", "PGK1", "PGAM1", "ENO1", "ENO2",
            "PKM", "LDHA", "LDHB"
            )
    )),
    .(gene_name, mRNA_by_VHL_loss, mTOR_inh_noVHL, mRNA_log2fc, mRNA_padj)
]


```


# Analysis of HIF dependent transcriptional regulation and mTOR dependent translational regulation


```{r VHL dependent mRNA changes and mTOR dependent translation changes}

print("Sanity check")
nrow(th.all.de.dte.res.dt[duplicated(gene_id)])

ggplot(
    th.all.de.dte.res.dt,
    aes(
        x = log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
        y = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
        color = VHL_target_RCC4
    )
) +
    geom_hline(yintercept = 0, color = "gray40") +
    geom_vline(xintercept = 0, color = "gray40") +
    scale_color_manual(values = c(
                           "VHL_loss_induced" = "firebrick",
                           "non_VHL_target" = "black",
                           "VHL_loss_repressed" = "black"
                       )) +
    geom_point() +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    ) +
    xlim(c(-4, 12)) + ylim(c(-1.5, 1)) +
    xlab("mRNA log2 fold change upon VHL loss in RCC4") +
    ylab("MRL log2 fold change with Torin 1 in RCC4")

hif.related.genes <- c(
    "HIF1A", "EPAS1", "ARNT",
    "EGLN1", "EGLN2", "EGLN3", "HIF1AN"   
)

ggplot(
    th.all.de.dte.res.dt,
    aes(
        x = log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
        y = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA
    )
) +
    geom_hline(yintercept = 0, color = "gray40") +
    geom_vline(xintercept = 0, color = "gray40") +
    scale_color_manual(values = c(
                           "VHL_loss_induced" = "firebrick",
                           "non_VHL_target" = "black",
                           "VHL_loss_repressed" = "black"
                       )) +
    geom_point(shape = 16) +
    gghighlight::gghighlight(
                     gene_name %in% hif.related.genes,
                     label_key = gene_name
                 ) +
    theme(
        legend.position = "bottom",
        aspect.ratio = 1
    ) +
    xlim(c(-4, 12)) + ylim(c(-1.5, 1)) +
    xlab("mRNA log2 fold change upon VHL loss in RCC4") +
    ylab("MRL log2 fold change with Torin 1 in RCC4")


th.all.de.dte.res.dt %$%
    cor.test(
        x = log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
        y = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
        method = "spearman",
        alternative = "two.sided"
    )

print(paste0(
    "The number of samples for this analysis was: ",
    nrow(th.all.de.dte.res.dt[
        !is.na(MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA)
    ])
))

```


```{r export table for publication}

trsl.change.by.class.dt <- copy(th.all.de.dte.res.dt[, .(
    gene_id, gene_name,
    mRNA_trsl_intersection_by_functions,
    VHL_target_RCC4,
    log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
    ## padj_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
    MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA
)])[
    mRNA_trsl_intersection_by_functions != "Others" &
    VHL_target_RCC4 == "VHL_loss_induced"
][, VHL_target_RCC4 := NULL][
    order(
        mRNA_trsl_intersection_by_functions,,
        - log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL
    )
]

setnames(
    trsl.change.by.class.dt,
    old = c(
        "mRNA_trsl_intersection_by_functions", 
        "log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL",
        "MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA"
    ),
    new = c(
        "Functional_classes",
        "mRNA_log2FC_upon_VHL_loss",
        "MRL_log2FC_with_Torin_1"
    )
)

fwrite(
    trsl.change.by.class.dt,
    file.path(s9.dir, "mTOR_sensitivity_of_HIF_targets_by_class.csv")    
)

```



# Differential sensitivity to mTOR inhibition of HIF-target mRNAs by their functional classes and HIFA isoforms


## Analysis of mTOR sensitivity of HIF-target mRNAs by their functional classes


```{r analysis of mrna vs translation by protein functions}

unhighlighted.color <- "gray60" #"#BEBEBEB2"

go.annot.colors <- c(
    "Glycolysis" = "#E88CA0",
    "Angiogenesis or vascular process" = "#008A42",
    "Others" = unhighlighted.color
)

## Density plot version
ggplot(
    data = th.all.de.dte.res.dt[VHL_target_RCC4 == "VHL_loss_induced"],
    aes(
        x = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
        color = forcats::fct_rev(mRNA_trsl_intersection_by_functions),
        fill = forcats::fct_rev(mRNA_trsl_intersection_by_functions)
    )
) +
    geom_density(
        alpha = 0.7,
        color = "black",
        bw = "SJ"
    ) +
    ## geom_rug(
    ##     data = th.all.de.dte.res.dt[
    ##         VHL_target_RCC4 == "VHL_loss_induced" &
    ##         mRNA_trsl_intersection_by_functions != "Others"
    ##     ],
    ##     size = 1
    ## ) +
    scale_fill_manual(values = go.annot.colors) +
    scale_color_manual(
        values = go.annot.colors,
        guide = "none"
    ) +
    guides(
        fill = guide_legend(reverse = TRUE, ncol = 1)
    ) +
    xlab("MRL log2 fold change with Torin 1") +
    ylab("Density of genes") +
    theme(
        legend.position = "bottom",
        legend.title = element_blank()
    )

print("The number of genes for this analysis")
th.all.de.dte.res.dt[
    VHL_target_RCC4 == "VHL_loss_induced" &
    !is.na(MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA),
    .N, by = mRNA_trsl_intersection_by_functions
]

print("in RCC4")
th.all.de.dte.res.dt[VHL_target_RCC4 == "VHL_loss_induced"] %$%
    pairwise.wilcox.test(
        x = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
        g = mRNA_trsl_intersection_by_functions,
        p.adjust.method = "none",
        alternative = "two.sided"
    )

## Just to confirm the robustness
print("in RCC4 VHL")
th.all.de.dte.res.dt[VHL_target_RCC4 == "VHL_loss_induced"] %$%
    pairwise.wilcox.test(
        x = MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA,
        g = mRNA_trsl_intersection_by_functions,
        p.adjust.method = "none",
        alternative = "two.sided"
    )

```


## Analysis of HIF2A/HIF1A binding ratio of HIF-target mRNAs by their functional classes


```{r hif2a hif1a binding ratio, fig.height = 7}

## Here I focus on the comparison of angiogenesis and vasucular process genes vs others
th.all.de.dte.res.dt[, `:=`(
    angiogenesis_or_others = case_when(
        mRNA_trsl_intersection_by_functions == "Angiogenesis or vascular process" ~
            "Angiogenesis or vascular process",
        TRUE ~ "Others"
    )
)]

## Load HIF binding data
tss.hif.pos.dt <- file.path(
    s7.2.hif1.2a.dir,
    "filtered-tss-and-nearest-hif-binding-position.csv"
) %>%
    fread

hif.induced.tss.hif.pos.dt <- tss.hif.pos.dt[
    tss_name %in%
    tss.de.res.dt[
        comparison_name == "RCC4_xx_HIF1B_N__noVHL_vs_VHL" &
        padj < 0.1 &  log2fc > 0
    ][order(shrlog2fc, decreasing = TRUE)][!duplicated(gene_id), tss_name]
]

hif.binding.and.trsl.dt <- merge(
    th.all.de.dte.res.dt,
    hif.induced.tss.hif.pos.dt[, .(gene_id, Fold, HIF2A_enrichment_with_dist_th)],
    by = "gene_id"
)

ggplot(
    data = hif.binding.and.trsl.dt[
        VHL_target_RCC4 == "VHL_loss_induced"
    ],
    aes(
        y = HIF2A_enrichment_with_dist_th,
        x = mRNA_trsl_intersection_by_functions
    )
) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_boxplot(outlier.shape = NA) +
    scale_x_discrete(guide = guide_axis(angle = 90)) +
    ggsci::scale_fill_jama() +
    ggsci::scale_color_jama() +
    theme(
        aspect.ratio = 3,
        legend.position = "bottom",
        legend.title = element_blank()
    ) +
    xlab("Functional class") +
    ylab("log2(HIF2A binding / HIF1A binding)")


stats.hif.binding.and.trsl.dt <- hif.binding.and.trsl.dt[
    !is.na(HIF2A_enrichment_with_dist_th) &
    VHL_target_RCC4 == "VHL_loss_induced"    
]

stats.hif.binding.and.trsl.dt[
  , .N, by = mRNA_trsl_intersection_by_functions
]

stats.hif.binding.and.trsl.dt %$%
    pairwise.wilcox.test(
        x = HIF2A_enrichment_with_dist_th,
        g = angiogenesis_or_others,
        p.adjust.method = "none",
        alternative = "two.sided"
    )

lapply(
    stats.hif.binding.and.trsl.dt[
        , levels(mRNA_trsl_intersection_by_functions)
    ],
    function(x){
        data.table(
            functional_class = x,
            wil_p = wilcox.test(
                x = stats.hif.binding.and.trsl.dt[
                    mRNA_trsl_intersection_by_functions == x,
                    HIF2A_enrichment_with_dist_th
                ],
                y = stats.hif.binding.and.trsl.dt[
                    mRNA_trsl_intersection_by_functions != x,
                    HIF2A_enrichment_with_dist_th
                ],
                alternative = "two.sided"
            )$p.value             
        )
    }
) %>%
    rbindlist %>%
    {.[functional_class != "Others"]} %>%
    {.[, padj := p.adjust(wil_p, method = "holm")]} %>%
    print

```


## Analysis of HIF2A-dependent induction of HIF-target mRNAs by their functional classes


```{r analysis of hif2a and vascular system remodling gene, fig.height = 7}

ggplot(
    data = th.all.de.dte.res.dt[
        VHL_target_RCC4 == "VHL_loss_induced" &
        !is.na(padj_786O_xx_HIF1B_N__noVHL_vs_VHL)
    ],
    aes(
        x = mRNA_trsl_intersection_by_functions,
        y = log2fc_786O_xx_HIF1B_N__noVHL_vs_VHL
    )
) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_boxplot(outlier.shape = NA) +
    scale_x_discrete(guide = guide_axis(angle = 90)) +
    coord_cartesian(ylim = c(-1.5, 3)) +
    theme(
        legend.position = "bottom",
        legend.title = element_blank(),
        aspect.ratio = 3
    ) +
    ylab("mRNA log2 fold change upon VHL loss in 786-O") +
    xlab("Functional class")

## For statisitical analysis of mRNA induction in 786-O
for.stat.th.all.de.dte.res.dt <- copy(
    th.all.de.dte.res.dt[
        VHL_target_RCC4 == "VHL_loss_induced",
        .(gene_id, gene_name,
          mRNA_trsl_intersection_by_functions,
          angiogenesis_or_others,
          translational_regulation_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
          padj_RCC4_xx_HIF1B_N__noVHL_vs_VHL, log2fc_RCC4_xx_HIF1B_N__noVHL_vs_VHL,
          padj_786O_xx_HIF1B_N__noVHL_vs_VHL, log2fc_786O_xx_HIF1B_N__noVHL_vs_VHL,
          meanNormCount_treated_786O_xx_HIF1B_N__noVHL_vs_VHL,
          meanNormCount_base_786O_xx_HIF1B_N__noVHL_vs_VHL)
    ]
)

for.stat.th.all.de.dte.res.dt[!is.na(padj_786O_xx_HIF1B_N__noVHL_vs_VHL)] %$%
    pairwise.wilcox.test(
        x = log2fc_786O_xx_HIF1B_N__noVHL_vs_VHL,
        g = angiogenesis_or_others,
        p.adjust.method = "none",
        alternative = "two.sided"
    )

## Show the number of samples, and the sample with missing value
for.stat.th.all.de.dte.res.dt[
    !is.na(padj_786O_xx_HIF1B_N__noVHL_vs_VHL),
    table(angiogenesis_or_others)
]

for.stat.th.all.de.dte.res.dt[
    is.na(padj_786O_xx_HIF1B_N__noVHL_vs_VHL),
    table(angiogenesis_or_others)
]


```


# Analysis of mTOR dependent translational regulation as a function of HIF pathway acivitity


```{r mTOR inhibition effect upon VHL loss}

th2.all.de.dte.res.dt <- th.all.de.dte.res.dt[
    gene_id %in% Reduce(
                     intersect,
                     list(
                         all.filtered.gene.dt[RCC4_noVHL_NA == TRUE, gene_id],
                         all.filtered.gene.dt[RCC4_noVHL_Torin1== TRUE, gene_id],
                         all.filtered.gene.dt[RCC4_VHL_NA == TRUE, gene_id],
                         all.filtered.gene.dt[RCC4_VHL_Torin1== TRUE, gene_id]
                     ))
]

ggplot(
    data = th2.all.de.dte.res.dt,
    aes(
        x = MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA,
        y = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA
    )
) +
    geom_abline(intercept = 0, slope = 1, color = "gray60") +
    geom_point() +
    geom_smooth(method = "lm") +
    coord_cartesian(xlim = c(-1.5, 0.5), ylim = c(-1.5, 0.5)) +
    xlab("MRL log2 fold change with Torin 1 in RCC4 VHL") +
    ylab("MRL log2 fold change with Torin 1 in RCC4") +
    theme(
        aspect.ratio = 1
    )

th2.all.de.dte.res.dt %$%
    cor.test(
        x = MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA,
        y = MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA,
        method = "pearson",
        alternative = "two.sided"
    )

print(
    paste0(
        "The number of genes analysed was: ",
        nrow(
            th2.all.de.dte.res.dt[, .(
            MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA,
            MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA
            ) %>% {.[complete.cases(.)]}]
        )
    )
)

th2.all.de.dte.res.dt %$%
    lm(
        MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA ~
            MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA,
    ) %>% summary

## Additional confirmation of the offset from the line of identity
th2.all.de.dte.res.dt %$%
    lm(
        MRL_log2fc_RCC4_noVHL_EIF4E2_yy_xx__Torin1_vs_NA ~
            MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA +
            offset(MRL_log2fc_RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA),
    ) %>% summary


```



# Session information


```{r sessionInfo}

sessionInfo()


```
