---
title: "s10-1 Validation of HP5 by RT-qPCR"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
     fig_width: 5
     fig_height: 5
---


# Overview

The ability of HP5 to resolve mRNA isoforms will be evaluated in this script.
For this purpose, the polysome distribution of selected mRNA isoforms estimated by the HP5 data (the average of three clones of RCC4 VHL) was compared with that estimated by RT-qPCR (RCC4 VHL clone 1).


# Environment setup


```{r load_libraries, message = FALSE, warning = FALSE}

temp <- sapply(list.files("../functions", full.names = TRUE), source)
temp <- sapply(list.files("./functions", full.names = TRUE), source, chdir = TRUE)

library("readxl")

processors <- 8

set.seed(0)

```


```{r define_directory}

## sample.file <- file.path("../../data/sample_data/processed_sample_file.csv")
annot.dir <- normalizePath(file.path("../../annotation/"))
annot.ps.dir <- file.path(annot.dir, "hg38_annotation/processed_data/")
annot.R.file <- list.files(
    annot.ps.dir,
    pattern = glob2rx("*primary_transcript_annotation*.rdata"),
    full.names = TRUE
)
load(annot.R.file)

results.dir <- file.path("../../results")

s8.dir <- file.path(results.dir, "s8-analysis-of-translation")
s8.1.dir <- file.path(s8.dir, "s8-1-differentially-translated-mRNAs")
s8.1.2.dir <- file.path(s8.1.dir, "tx-level-dte")

create.dirs(c(
))

## Load differential expression analysis
source(file.path("../s6-differential-expression-and-tss-usage/functions/load_total_analysis_results.R"), chdir = TRUE)

```

# Load and preprocess HP5 and 5' end-Seq data

## RNA standard normalised data

```{r read_start_count_data}

norm.count.dt <- fread(
    file.path(
        s8.1.2.dir,
        "RCC4_VHL_EIF4E2_yy_xx__Torin1_vs_NA-normalized_count.csv"
    )
)

```

## RNA expression data

```{r load_expression_data}

rcc4.tss.de.res.dt <- tss.de.res.dt[
    comparison_name == "RCC4_xx_HIF1B_N__noVHL_vs_VHL"
]

rcc4.tss.de.res.dt[
  , meanNormCount_base_gene_max := max(meanNormCount_base, na.rm = TRUE),
    by = gene_id
]

## Only isoforms expressed more than 10% of the most highly expressed isoform were analysed
rcc4.tss.de.res.dt <- rcc4.tss.de.res.dt[
    meanNormCount_base > meanNormCount_base_gene_max * 0.1
]

```

# Analysis of polysome distribution of the selected mRNA isoforms using the HP5 data


```{r plot_hp5_results}

sl.genes <- c("DEK", "EIF5A", "RBCK1", "TPGS2")

print("The following genes and isoforms will be analysed:")
rcc4.tss.de.res.dt[
    gene_name %in% sl.genes, .(gene_id, gene_name, tss_name)
]

tss.per.gene.dts <- list(
    DEK = data.table(
        gene_name = "DEK",
        tss_class = c("dist", "proxy"),
        tss_name = c("ENSG00000124795_1", "ENSG00000124795_2")
    ),
    EIF5A = data.table(
        gene_name = "EIF5A",
        tss_class = c("dist", "dist", "proxy"),
        tss_name = c(
            "ENSG00000132507_1",
            "ENSG00000132507_2",
            "ENSG00000132507_3"
        )
    ),
    RBCK1 = data.table(
        gene_name = "RBCK1",
        tss_class = c("dist", "both", "both"),
        tss_name = c(
            "ENSG00000125826_2",
            "ENSG00000125826_3",
            "ENSG00000125826_4"
        )
    ),
    TPGS2 = data.table(
        gene_name = "TPGS2",
        tss_class = c("dist", "both"),
        tss_name = c("ENSG00000134779_1", "ENSG00000134779_2")
    )
)


```


```{r normalised_count_data_for_plotting}

sl.norm.count.dt <- merge(
    rbindlist(tss.per.gene.dts)[, .(tss_name, tss_class)],
    norm.count.dt,
    by = "tss_name"
)

sl.norm.count.dt[, tss_name := paste0(gene_name, "_", tss_class)]

## Collapse count data for multiple TSSs
sl.norm.count.dt <- sl.norm.count.dt[
  , lapply(.SD, sum, na.rm = TRUE), by = tss_name,
    .SDcols = grep("RCC4", colnames(sl.norm.count.dt)) 
]

calcNormCountStats <- function(norm.count.dt, ref.col = "tss_name", grep.col = "^RCC4"){

    m.norm.count.dt <- melt(
        norm.count.dt,
        id.vars = ref.col,
        measure.vars = grep(
            grep.col,
            colnames(norm.count.dt)
        ),
        variable.name = "sample_name",
        value.name = "norm_count"
    ) %>%
        {.[, `:=`(
             VHL = str_split_fixed(sample_name, "_", n = 8)[, 2],
             EIF4E2 = str_split_fixed(sample_name, "_", n = 8)[, 3],
             clone = str_split_fixed(sample_name, "_", n = 8)[, 5],
             treatment = str_split_fixed(sample_name, "_", n = 8)[, 6],
             fraction = str_split_fixed(sample_name, "_", n = 8)[, 7]
         )]}

    m.norm.count.dt[
      , norm_count_sum := sum(norm_count),
        by = list(get(ref.col), VHL, EIF4E2, clone, treatment) 
    ][, norm_count_ratio := norm_count / norm_count_sum]
    
    norm.count.summary.dt <- m.norm.count.dt[, list(
        norm_count_ratio_mean = mean(norm_count_ratio),
        norm_count_ratio_sd = sd(norm_count_ratio)
    ), by = list(get(ref.col), VHL, EIF4E2, treatment, fraction)]

    setnames(norm.count.summary.dt, old = "get", new = ref.col)
    
    return(norm.count.summary.dt)
}

norm.count.summary.dt <- calcNormCountStats(
    sl.norm.count.dt,
    ref.col = "tss_name",
    grep.col = "RCC4_VHL_EIF4E2_NA_[[:digit:]]_NA_ribo[[:digit:]]"
)

norm.count.summary.dt[, `:=`(
    gene = str_split_fixed(tss_name, "_", n = 2)[, 1],
    type = str_split_fixed(tss_name, "_", n = 2)[, 2],
    experiment = "HP5\n(average of 3 clones)"
)]

```

# Analysis of polysome distribution of the selected mRNA isoforms using RT-qPCR data


```{r RT_qPCR}

qpcr.res.file <- file.path("../../data/others/RT-qPCR_all-results.xlsx")

qpcr.res.dt <- rbind(
    read_excel(qpcr.res.file, sheet = "replicate_1") %>% data.table %>%
    {.[, replicate := 1]},
    read_excel(qpcr.res.file, sheet = "replicate_2") %>% data.table %>%
    {.[, replicate := 2]}
)[
  , c("fraction", "gene", "type", "Ct", "replicate"), with = FALSE
][!is.na(fraction)]

qpcr.standard.dt <- rbind(
    read_excel(qpcr.res.file, sheet = "standard_1") %>% data.table,
    read_excel(qpcr.res.file, sheet = "standard_2") %>% data.table
)[, c("dilution", "gene", "type", "Ct"), with = FALSE]

qpcr.standard.summary.dt <- qpcr.standard.dt[, list(
    Ct_mean = mean(Ct),
    Ct_sd = sd(Ct)
), by = list(gene, type, dilution)]

qpcr.primer.eff.dt <- qpcr.standard.summary.dt[, as.list(
    coef(lm(Ct_mean ~ log10(1/dilution)))
), by = list(gene, type)][
  , primer_eff := 10^(-1/ get("log10(1/dilution)")) - 1
][order(gene == "ERCC-00002", gene, type)]

print(qpcr.primer.eff.dt)

print("Rsquared values for confirmation:")
qpcr.standard.summary.dt[, list(
    rsq = summary(lm(Ct_mean ~ log10(1/dilution)))$r.squared
), by = list(gene, type)][order(gene == "ERCC-00002", gene, type)]

```


```{r qpcr-result-analysis}

quant.dt <- merge(
    qpcr.res.dt,
    qpcr.primer.eff.dt[, .(gene, type, primer_eff)],
    by = c("gene", "type")
)

ercc.quant.dt <- quant.dt[gene == "ERCC-00002"]
ercc.quant.dt[, `:=`(
    ercc_input = (2 * primer_eff)^(-Ct)
)]

gene.quant.dt <- quant.dt[gene != "ERCC-00002"]
gene.quant.dt[, `:=`(
    gene_input = (2 * primer_eff)^(-Ct)
)]

all.quant.dt <- merge(
    gene.quant.dt,
    ercc.quant.dt[
      , c("replicate", "fraction", "ercc_input"), with = FALSE
    ],
    by = c("replicate", "fraction")
)

all.quant.dt[, `:=`(
    norm_input = gene_input / ercc_input
)]
all.quant.dt[
  , norm_input_sum := sum(norm_input), by = list(replicate, gene, type)
]
all.quant.dt[
    , norm_input_ratio := norm_input / norm_input_sum
]

qpcr.all.quant.res.dt <- all.quant.dt[, list(
    norm_count_ratio_mean = mean(norm_input_ratio),
    norm_count_ratio_sd = sd(norm_input_ratio),
    experiment = "RT-qPCR\n(clone 1)"
), by = list(gene, type, fraction)]

```

# Comparison of the HP5 data and RT-qPCR


```{r comparison_HP5_and_RT-qPCR, fig.height = 5, fig.width = 7}

all.res.dt <- rbind(
    norm.count.summary.dt[, colnames(qpcr.all.quant.res.dt), with = FALSE],
    qpcr.all.quant.res.dt
)

all.res.dt[
  , fraction := case_when(
        fraction == "ribo8" ~ "8+",
        TRUE ~ gsub("ribo", "", fraction)
    )
]

ggplot(
    data = all.res.dt,
    aes(
        x = fraction,
        y = norm_count_ratio_mean,
        ymin = norm_count_ratio_mean - norm_count_ratio_sd,
        ymax = norm_count_ratio_mean + norm_count_ratio_sd,
        color = type,
        fill = type,
        group = type
    )
) +
    geom_ribbon(alpha = 0.3, color = NA) +
    geom_line(size = 1) +
    facet_grid(experiment ~ gene) +
    theme_bw() +
    theme(
        legend.position = "bottom",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    scale_x_discrete(guide = guide_axis(angle = 90)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_color_bright() +
    scale_fill_bright() +
    ylab("Relative mRNA abundance") +
    xlab("Polysome fraction")

```



# Session information

```{r sessionInfo}

sessionInfo()

```
